### 2016-07-06 doc/publications/plos16/plos16.tex
changes to plos16.tex
added to .bib
changed layout of page
changes text

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

@@

\reversemarginpar

@@

\section*{Notes}
\begin{itemize}
\item what are you guys editing this with?  what about some carriage
  returns or line feeds or something!
\item Are we going to use the TCGA data in this paper?
\item are we going to run sparkML or Cursed forest on the extended
  simulation data \cite[]{Genuer.et.al.2010}? Has Aidan done this already?
\end{itemize}

@@

While unsupervised clustering is useful for selected clinical applications~\cite{Li2015}, a more common analysis task is
supervised learning, i.e. given a phenotype of clinical label determine the most associated features.  One of the
simplest supervised machine learning methods is logistic regression (LR), which creates a decision boundary by fitting a
linear combination of the feature variables.  Processing the `wide' genomic data poses difficulties for machine learning
applications as having more features $p$ than samples $n$ causes the model to overfit the data and generalize poorly.
This is known as the `curse of dimensionality'~\cite{Bauer2014}.  Random Forests~\cite{Breiman.2001} (RF) as an ensemble of weak learners 
\marginpar{weak learners only in a misuse of the term. ``weak learners'' was applied to decision stumps}
%http://stats.stackexchange.com/questions/77018/random-forest-is-it-a-boosting-algorithm
%http://stats.stackexchange.com/questions/173390/gradient-boosting-tree-vs-random-forest
overcomes this issue and is hence ideally suited to this application.  Furthermore, RF are able to capture interactions
between features, which is of importance for modelling complex polygenic diseases.

% We also investigated classifying the samples using Random Forests.  We chose random forests due to its ability to
% capture complex interactions between features. Also, due to it being an ensemble method where trees can be formed
% independently, which, is ideal for parallelisation. Furthermore, the ensemble of weak learners can help to minimise
% overfitting and variance.

@@

As discussed in the Introduction, random forests are robust against overfitting
\marginpar{strong claim -- see disussion in Segal 2004, perhaps}
and hence can generalise for datasets where $p > n$, however here we demonstrate that this holds true in extreme cases where $p \gg n$.

@@

@@

\subsection*{decision trees }
\begin{itemize}
\item benefits os a decision tree
  \begin{itemize}
  \item invariant to scale of the input data
  \item handles categorical and real valued inputs 
  \item can handle missing values 
  \item insensitive to useless predictors
  \end{itemize}
\item benefits of random forests
  \begin{itemize}
  \item variable importance 
  \item inherits the qualities of tree models
  \end{itemize}
\end{itemize}

>>>>

\documentclass[10pt,a4paper]{article}  %was \documentclass[10pt,letterpaper]{article}

\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

@@

\let\oldmarginpar\marginpar
\renewcommand\marginpar[1]{\-\oldmarginpar[\raggedleft\footnotesize #1]%hv
{\raggedright\footnotesize #1}}
\reversemarginpar
\setlength{\marginparwidth}{5cm}

@@

\section*{Notes}
\begin{itemize}
\item what are you guys editing this with?  what about some carriage
  returns or line feeds or something!
\item Are we going to use the TCGA data in this paper?
\item are we going to run sparkML or Cursed forest on the extended
  simulation data \cite[]{Genuer.et.al.2010}? Has Aidan done this already?
\item I dont think I can get R to handle chromosme 1 etc. May have to run CF on chromosme 22, see table
  \ref{table:smaller_data}
\item check Piotr's simulation
\item run an argument as follows:
  \begin{itemize}
  \item variable  selection may succeed where accurate prediction does not
  \item consider  \cite{Segal.2004} examples
  \item  perhaps we argue that in the case where the relationship of $X$ and $y$ takes a particular form we can recover
    the reltionship for very large $p$p
  \end{itemize}
\item \cite{Tuv.et.al.2009} feature selection via adding permuted variables. Do we want to consider this?
\item chek out varSelRF
\end{itemize}

@@

While unsupervised clustering is useful for selected clinical applications~\cite{Li2015}, a more common analysis task is
supervised learning, i.e. given a phenotype of clinical label determine the most associated features.  One of the
simplest supervised machine learning methods is logistic regression (LR), which creates a decision boundary by fitting a
linear combination of the feature variables.  Processing the `wide' genomic data poses difficulties for machine learning
applications as having more features $p$ than samples $n$ causes the model to overfit the data and generalize poorly.
This is known as the `curse of dimensionality'~\cite{Bauer2014}.  
\marginpar{is this the reference we want for the `curse of dimensionality'}

\subsection*{RF}

Random Forests~\cite{Breiman.2001} (RF) apply the technique of bootstrap aggregation or bagging to decision trees.  The training
data is independently drawn from the joint distribution of $(X,Y)$ and comprises $n$ $(p+1)$-tuples $(x_1,y_1),\ldots, (x_n,y_n)$.
$X$ and $\theta_b$ for $b=1,\ldots,B$ are i.i.d. random vectors and a tree is grown on each of these $\theta_b$  samples 
($X$ is randomly sampled $B$ times with replacement and a tree is grown on each of these $B$ samples).

The results are combined in an appropriate way, i.e. for a regression problem,
\begin{equation*}
{\bar {h}}={\frac {1}{B}}\sum _{b=1}^{B} h(x;\theta_b).
\end{equation*}
%where $x$ is the covariate vector of length $p$
%with associated random vector $X$ and the $\theta_b$ are i.i.d. random vectors.

The average prediction error for an individual tree $h(X; \theta)$ is
\begin{equation}
PE_t = E_\theta E_{X,Y} (Y-h(X; \theta))^2.
\end{equation}
Assume that, for all  $\theta$, the tree is unbiased, i.e., $EY= E_X h(X; \theta)$. Then
\begin{equation}
PE_f \leq \rho PE_t
\end{equation}
where $\rho$ is the weighted correlation between residuals $(Y-h(X;\theta))$ and $(Y-h(X;\theta^\prime))$ for independent $\theta,
\theta^\prime$.  
So what is required for  accurate random forest regression is (i) low correlation between residuals of differing tree in
the forest, and (ii) low prediction error for the individual trees \cite{Segal.2004}.

\subsection*{benefits of RF}

\marginpar{weak learners only in a misuse of the term. ``weak learners'' was applied to decision stumps}
%http://stats.stackexchange.com/questions/77018/random-forest-is-it-a-boosting-algorithm
%http://stats.stackexchange.com/questions/173390/gradient-boosting-tree-vs-random-forest
%overcomes this issue and is hence ideally suited to this application.  Furthermore, RF are able to capture interactions
%between features, which is of importance for modelling complex polygenic diseases.

% We also investigated classifying the samples using Random Forests.  We chose random forests due to its ability to
% capture complex interactions between features. Also, due to it being an ensemble method where trees can be formed
% independently, which, is ideal for parallelisation. Furthermore, the ensemble of weak learners can help to minimise
% overfitting and variance.
Decision trees have a number of desirable features, they:
\begin{itemize}
  \item are scale invariant;
  \item can handle categorical and real valued inputs;
  \item can handle missing values;
  \item are insensitive to useless predictors;
  \item are able to capture interactions between features (which is of importance for modelling complex polygenic diseases).
  \end{itemize}
However, the tree fitting algorithm is greedy and may generate an unstable model. That is, a small change in the data may
lead to a very different model. 
\marginpar{a potential problem  with tree models is the tendencey to split on categorical variables with a high number of
  levels, even when better splits are available}
Random forests are a variance reduction technique. They take a tree models which have a low bias but a high variance and
by combining them reduce the variance. The prediction error is the sum of the variance and the bias squared.

The benefits of random forests models are
  \begin{itemize}
  \item they give an out-of-bag estimate of model accurace;
  \item they provide a measure of variable importance;
  \item they inherit many of the qualities of tree models.
\marginpar{but not the handling of missing velues via surrogate variables. This would be difficult as the trees as split
  on random subsets of variables}
  \end{itemize}

@@

As discussed in the Introduction,
\marginpar{I think this has been removed from the introduction}
random forests are resistent against overfitting.
However it is not true that they will not overfit; see \cite{Segal.2004} where examples of RF overfitting are given 
\marginpar{discuss these examples}

Subject to certain conditions, RF will not overfit even in the  $p > n$ case. 
Here we demonstrate that this may hold true in extreme cases where $p \gg n$.

@@

\begin{table}
\begin{minipage}{\textwidth}
\begin{center}
\begin{tabular}{|p{4.2cm}|l|l|l|l|l|l|l|}
\hline
\bf{Genome }  & \bf{method} & \bf{Accuracy} & \bf{Runtime} & \bf{Memory/Node} \\
\hline
 chr 22 - phase1 (2.1GB)  &  RF (R)\footnote{10 nodes, 500 trees, dpas03} 
                                   &  0.456 & 2.89 & 18GB \\ \hline
\end{tabular}
\medskip
\begin{flushleft} 
\caption{some R results on smaller data sets.}
\label{table:smaller_data}
\end{flushleft}
\end{center}
%\end{sidewaystable}
\end{minipage}
\end{table}

% \begin
% \begin{table}[!ht]
% \caption{
% {\bf Performance comparison between the different Machine Learning algorithms.}}
% \begin{tabular}{|p{4.2cm}|l|l|l|l|l|l|l|}
% \hline
% \bf{Genome }  & \bf{method} & \bf{Accuracy} & \bf{Runtime} & \bf{Memory/Node} \\
% \hline
%  chr 22 - phase1 (2.1GB)  &  RF (R) &  0.456 & 2.89 & 18GB \\ \hline
% \end{tabular}
% \begin{flushleft} 
% \end{flushleft}
% \label{table1}
% %\end{adjustwidth}
% \end{table}

@@
